---
title: "miQTL"
format: html
editor: visual
editor_options: 
chunk_output_type: console
---

## miQTL

sparcc data used to simulate phenotypes from a mulitpartent population (DSPR / CC) sparc cache repo is needed to ensure the simulated data is representing the CC

```{r}

#General setup and package installation

remotes::install_github("gkeele/miqtl")
devtools::install_github("gkeele/sparcc")
install.packages("data.table")
library("sparcc")
library("miqtl")
library(tidyverse)
library(dplyr)

```

Goal: run a genome scan of chrom 19 , transcript Cyp2c39

```{r}
# General data acquisition
genomecache <- ("~/ValdarLab/segments_happy_format_mm10")
allRNA <- data.table::fread(file = "../ValdarLab/allRNA.txt")

# format needed to run the genome scan
reader <- DiploprobReader(dataDir = genomecache)

data <- data.frame("Y" = allRNA$Cyp2c39, "Strain" = allRNA$STRAIN)

# h2lmm = heritability 2, linear mixed model
scan <- miqtl::scan.h2lmm(genomecache = genomecache, 
                    data = data, 
                    pheno.id = "Strain", 
                    geno.id = "Strain",
                    formula = Y ~ 1,
                    return.allele.effects = TRUE)

miqtl::genome.plotter.whole(list(scan))

miqtl::genome.plotter.chr(scan, chr = 19)

miqtl::allele.plotter.region(scan.object = scan, chr = 19)


```

# Now doing the scan without the function

1. Get the first locus at chrom 19

```{r}
locus <- reader$getFirstLocus(chrom = 19)

locus_matrix <- reader$getLocusMatrix(locus = locus, subjects = allRNA$STRAIN, model = "additive")

fit <- lm(data$Y ~ 1+locus_matrix )
fit0 <- lm(data$Y ~ 1 )

anova(fit, fit0)
pval <- anova(fit,fit0)$`Pr(>F)`[2]

scan$p.value[locus]

# sanity check, make sure the previous sig points are still showing as sig in the manual version
# can replace locus with locus_sig and then re-run. The new pvalue should be hella sig

locus_sig <- miqtl::grab.locus.from.scan(scan.object = scan)

```



```{r}

#tibble to store the anova comparison for the 2 fit models in the loop
sigRNA <- tibble(
  "locus" = character(length = length(loci)),
  "anova_pval" = numeric(length = length(loci)),
  "begin" = numeric(length = length(loci)),
  "end" = numeric(length = length(loci)),
)


## get locus range in reader for each locus in the loop. add that to the sigRNA and then re-plot

fit0 <- lm(allRNA$Cyp2c39 ~ 1, data = allRNA)
loci <- reader$getLoci(chrom = 19)

# for-loop made with Valdar; looping through each loci on the 19th chrom. 
for (ii in 1:length(loci)){
  print(loci[ii])
  locus_matrix <- reader$getLocusMatrix(locus = loci[ii], subjects = allRNA$STRAIN, model = "additive")
  fit1 <- lm(allRNA$Cyp2c39 ~ 1 + locus_matrix, data = allRNA)
  sum_model <- summary(fit1)
  result <- anova(fit0, fit1)
  
  sigRNA[ii, "anova_pval"] <- anova(fit1, fit0)$`Pr(>F)`[2]
  sigRNA[ii, "locus"] <- print(loci[ii])
  sigRNA[ii, "begin"] <- reader$getLocusRange(loci = loci[ii], scale = "Mb")[1]
  sigRNA[ii, "end"] <- reader$getLocusRange(loci = loci[ii], scale = "Mb")[2]
  
  
}

-----------------------------------------------------------------------------------


sigRNA_ABC <- tibble(
  "locus" = character(length = length(loci)),
  "anova_pval" = numeric(length = length(loci)),
  "begin" = numeric(length = length(loci)),
  "end" = numeric(length = length(loci))
)


fit0 <- lm(allRNA$Abcf1 ~ 1, data = allRNA)
loci <- reader$getLoci(chrom = 19)


for (ii in 1:length(loci)){
  print(loci[ii])
  locus_matrix <- reader$getLocusMatrix(locus = loci[ii], subjects = allRNA$STRAIN, model = "additive")
  fit1 <- lm(allRNA$Abcf1 ~ 1 + locus_matrix, data = allRNA)
  sum_model <- summary(fit1)
  result <- anova(fit0, fit1)
  
  sigRNA_ABC[ii, "anova_pval"] <- anova(fit1, fit0)$`Pr(>F)`[2]
  sigRNA_ABC[ii, "locus"] <- print(loci[ii])
  sigRNA_ABC[ii, "begin"] <- reader$getLocusRange(loci = loci[ii], scale = "Mb")[1]
  sigRNA_ABC[ii, "end"] <- reader$getLocusRange(loci = loci[ii], scale = "Mb")[2]
  
}


----------------------------------------------------------------------------------------


```



```{r}


sigRNA <- sigRNA %>% 
  mutate(locus_number = seq(243, by = 1, from = 1 ),
         log_p_cyp = -log10(anova_pval))
sigRNA |> 
  ggplot(aes(begin, log_p_cyp))+
  geom_line(color = "red")+
  labs(title = "y ~ 1+ Locus_Matrix", 
       x = "Position(Mb)",
       y = expression("-logP"[10]))+
  theme_minimal()


# same as above but for the ABC data
sigRNA_ABC <- sigRNA_ABC |> 
  mutate(locus_number = seq(243, by = 1, from = 1 ),
         log_p_ABC = -log10(anova_pval))


merged_df <- merge(sigRNA, sigRNA_ABC, by = "locus")

ggplot(merged_df) +
  geom_line(aes(x = begin.x, y = log_p_cyp, color = "Cyp_c23"), size = 0.7) +
  geom_line(aes(x = begin.y, y = log_p_ABC, color = "ABC"), size = 0.7) +
  geom_vline(xintercept = 38.2, col = "black", lwd = 0.5, linetype = "dashed") +
  scale_color_manual(name = "Transcripts", values = c("Cyp_c23" = "red", "ABC" = "blue")) +
  xlab("Position (Mb)") +
  ylab("-Log10 P-value") +
  theme_minimal()


  
# Finding the most significant locus for the red line (cypc23)
min_locus <- which.min(merged_df$anova_pval.x)
print(merged_df[139,])

#locus of interest
UNC30286399

locus_matrix <- reader$getLocusMatrix(locus = "UNC30286399", subjects = allRNA$STRAIN, model = "additive")

fit1 <- lm(allRNA$Cyp2c39 ~ +1 + locus_matrix, data = allRNA)
fit2 <- lm(allRNA$Cyp2c39 ~ -1 + locus_matrix, data = allRNA) 

#saving betas 
fit1df <- fit1$coefficients
fit2df <- fit2$coefficients

# running into problems with the vectors being the wrong length so skipping and going it the manual way

model_1 <-c( 0.3158466,0.5264983,  -1.4554181,  0.4314216, 0.4164416, -0.5145771, -0.7076691, -0.8455184, NA)
model_2 <- c( NA, 0.8423450,-1.1395714, 0.7472682,  0.7322882, -0.1987305,  -0.3918225, -0.5296717, -0.5296717)
haplotypes <- c("intercept", "A.J", "C57B1.6J","X129SvlmJ", "NOD.LtJ", "NZO.HlLtJ", "CAST.EiJ", "PWK.PhJ", "WSB.EiJ")

model_fits <- data.frame(intercept_fit = model_1, no_intercept_fit = model_2, betas = haplotypes)

model_fits <- model_fits %>% 
  mutate(Position_MB = seq(9, by = 1, from = 1 ))

model_fits |>
  ggplot(aes(Position_MB, intercept_fit))+
  geom_line(color = "blue")+
  geom_line(aes(Position_MB, no_intercept_fit), color = "red")

# model comparison
anova(fit1, fit2)




```

Multi-allelic Investigation
```{r}


devtools::install_github("wesleycrouse/TIMBR", build_vignettes=T)
library(TIMBR)
vignette("use_TIMBR")

#prior info on the number of models we have aka nothing 
# itll be an 8xK matrix, but now were looking at one allele so our K = 1

# Y ( first param for TIMBR) should be a list of the expression data from before (phenotype)
exp <- allRNA$Cyp2c39

#changing the format so we can edit in the correct format instead of converting later
prior.D <- mcv.data$prior.D
# setting the probability matrix we've created
prior.D$P <- mm_timber
#prior.D[[A]] is a basic probability matrix for heterogeneity,  homos and hets 

# saying there are no linked alleles, 8 different options
prior.M = list(model.type = "fixed", M.IDs = "1", "2","3","4","5","6","7","8")

# prior prob matrix for the distribution, similar to the past haplos but full creates diplos 
mm_timber <- reader$getLocusMatrix(locus = locus_sig, subjects = allRNA$STRAIN, model = "full")


#running timber
results1 <-  TIMBR(y = exp, prior.M = list(model.type = "fixed", M.IDs = c("1", "2","3","4","5","6","7","8")), 
      prior.D = prior.D
)

# results are going to update the priors/ probabilities of that general matrix we put forth in the prior D Matrix
results1$p.D.given.y


colMeans(results1$post.hap.effects)

TIMBR.plot.haplotypes(results1)
as.matrix(head(results1$p.M.given.y, 10))



_______________

# adding allele pairs, we see shared phenotype and haplos based on the pic above. tri-allele check
prior.M.biallelic <- list(model.type="fixed", M.IDs = "0,1,0,0,2,2,2,2")
results.biallelic <- TIMBR(exp, prior.D, prior.M.biallelic, verbose=F)
as.matrix(head(results.biallelic$p.M.given.y, 10))

results.biallelic$post.hap.effects[1:10,]

#overlay single allele with tri-alleleic.
TIMBR.plot.haplotypes(results.biallelic, TIMBR.output.bkgrd = results1)


#trying a uniform distribution
prior.M.uniform <- results1$prior.M$uniform
prior.M.uniform <- list(model.type="fixed", M.IDs = "0,0,0,0,0,0,0,0")
results.uniform <- TIMBR(exp, prior.D, prior.M.uniform, verbose=F)

TIMBR.plot.haplotypes(results.uniform, TIMBR.output.bkgrd = results1)

_________________

#Modeling the allele series with a Chinese restaurant process prior distribution

#filling sub-setted values withing prior.M 
results1$prior.M$model.type <- "crp"
results1$prior.M$prior.alpha.type <- "gamma"
results1$prior.M$prior.alpha.shape <- 1
results1$prior.M$prior.alpha.rate <- 2.33

#replacing prior.M with new values
prior.M.crp <- results1$prior.M
str(prior.M.crp)


results.crp <- TIMBR(exp, prior.D, prior.M.crp, verbose = F)
as.matrix(head(results.crp$p.M.given.y, 10))

TIMBR.plot.haplotypes(results.crp, TIMBR.output.bkgrd = results1)

```

Next Step : Visualizing prior distribution of the # of alleles


See: TIMBR Vignettes : USE TIMBR

Data General: 
Control mice to compare to a drugged mice group. Characterizing the baseline genetic variation for drug metabolizing proteins in the livers of the CC. 

Next Step: 
pulling out the posterior odds of the models to see what model is being preferred
ie, which model fits better, 2haps 3haps 8haps etc.


#Computing the prior distribution with Ewens’s sampling formula
```{r}

library(ape)
trees <- results.crp$trees
plot.phylo(trees[[1]])
#error, found fewer than 2 tips in the tree


#prior.alpha <- list(type="gamma", shape=1, rate=2.333415)
#prior.M.tree.1 <- ewenss.calc(trees[[1]], prior.alpha)
#str(prior.M.tree.1)


#trees is an object that results1 doesn't have so this wont work
prior.M.tree <- results1$prior.M$tree
str(prior.M.tree)



```


#Visualizing the prior distribution of number of alleles


```{r}


prior.alpha <- list(type="gamma", shape=1, rate=2.333415)
# if wanting to compare the other models etc. 
# prior.alpha <- list(type="fixed", alpha=1)

prior.M.crp.list <- ewenss.calc(8, prior.alpha)
number.alleles <- sapply(prior.M.crp.list$M.IDs, 
                         function(x){max(as.numeric(unlist(strsplit(x, split=","))))+1})
prior.number.alleles.uniform <- table(number.alleles) / length(number.alleles)
prior.number.alleles.crp <- exp(tapply(prior.M.crp.list$ln.probs, number.alleles, 
                                       matrixStats::logSumExp))

plot(1:8, prior.number.alleles.crp, type="l", lty=5, col="white", 
     ylim=c(0,1), ylab="Prior Probability", xlab="Number of Alleles", 
     main="Prior Distribution of Alleles")
lines(1:8, prior.number.alleles.crp, lty=5, col="orange")
points(1:8, prior.number.alleles.crp, pch=22, bg="orange")
lines(1:8, prior.number.alleles.uniform, lty=5, col="darkcyan")
points(1:8, prior.number.alleles.uniform, pch=22, bg="darkcyan")

legend(1, legend=c("Uniform", "CRP"), col=c("darkcyan", "orange"), lty=5)






```































